On Sample Size for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2023-04-26 Manuscript/manuscript.tex
comments through discussion of Figure 1

\bigskip
\noindent{\sc Keywords}:
dependent data; gsimulation
\end{abstract}

@@

The necessary sample size for bootstrap standard errors to provide valid
uncertainty measures in practice has not been extensively studied. For
independent data and non-block bootstrap, \citet{hesterberg2015teachers} notes
that while percentile-based CIs from nonparametric bootstrap
are more accurate than $t$-intervals for larger sample sizes, they are
less accurate for smaller sample sizes. In the contect of structural equation
modeling, \citet{nevitt2001performance} find
that a sample size of 200--1000 is usually sufficient for interval estimation
using standard nonparametric bootstrap. For dependent data, the necessary sample
size for block bootstrap is even less clear. In a linear regression setting with
dependent data, \citet{goncalves2005bootstrap} find that, standard error
estimates from block bootstrap in small samples may be more accurate than
inference from closed-form asymptotic estimates, but block bootstrap percentile
CIs still do not have sufficient coverage even for sample size
$n = 1024$.

@@

The remainder of the papers is organized as follows.
Section~\ref{sec:bbci} reviews block bootstrap procedures and how to use block
bootstrap estimates construct CIs. Section~\ref{sec:simu} reports a simulation
study comparing the coverage rates of the block bootstrap CIs. A discussion
concludes in Section~\ref{sec:disc}.

@@

Consider a stationary time series $\{X_t: t = 1, \ldots, n\}$ with length~$n$.
Our goal is to construct a CIs for a parameter $\theta$ in the
data generating model of the series. Suppose that $\hat\theta_n$ is a point
estimator of~$\theta$ based on the observed series. Bootstrap is a powerful
approach to construct CIs. If the observations in the series
were independent, a standard nonparametric bootstrap procedure would draw a
large number~$B$ bootstrap copies of the observed data, and calculate an
estimate $\hat\theta_n^{(b)}$ for each copy $b = 1, \ldots, B$. The uncertainty
of $\hat\theta_n$ is then estimated by the empirical uncertainty of the
$\hat\theta_n^{(b)}$'s. When serial dependence is present, the bootstrap
procedure needs to preserve the serial dependence. Block bootstrap was
motivated for this situation.

@@

Block bootstrap preserves the serial dependence in the observed data by
partitioning the data into blocks and perform bootstrap on the blocks.
In particular, consider block size~$l$ and, for convenience, suppose that
$n$ is a multiple of $l$ such that there are $k = n / l$ blocks. Each block~$j$
is $Y_j = \{X_{(j - 1) * l + 1}, \ldots, X_{(j - 1) * l + l}\}$,
$j = 1, \ldots,   k$.  Then, we sample $k$ blocks of $Y_j$'s from the set 
$\{Y_1, \ldots, Y_k\}$ with replacement and concatenate the $k$ sampled blocks
in the order they are picked to form a bootstrap sample of the data. The
formation of the bootstrap sample ensures that the between-block dependence is
week and that the within-block serial dependence is preserved. Because the
blocks here are non-overlapping, this bootstrap approach is known as
non-overlapping block bootstrap, or simple block bootstrap.

Alternatively, block-bootstrap can be done with overlapping or moving blocks.
Define moving blocks $Z_j = \{X_j, \ldots, X_{j + l - 1}\}$,
$j = 1, \ldots, n - l + 1$. Now we draw $k$ blocks from the $(n - l + 1)$
blocks
of $Z_j$'s with replacement and then aligning them in the order they were
picked
to form a block bootstrap sample. If $n$ is not a multiple of~$l$, the last
block selected will be reduced in size so that the final size of the
block bootstrap sample is $n$. It is also possible to implement moving block
bootstrap while allowing blocks to wrap around the end of the series. In other
words, define moving blocks (assuming $l > 1$):
\jy{Any equation must be a part of a sentence and should be punctuated.}
\jy{Recall a blank line means a new paragraph.}
\jy{Define your response command with color and respond}

@@

Suppose that we have repeated the steps in the last subsection $B$ times, and
that for $b \in \{1, \ldots, B\}$, we have obtained an estimator
$\hat\theta_n^{(b)}$ based on the $b$th bootstrap sample using the same method
that was applied to $\{X_t: t = 1, \ldots, n\}$ to obtain $\hat\theta_n$.
Now the question is, how to construct CIs for $\theta$
using the $B$ bootstrap point estimates
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$.
We consider three kinds of CIs.

@@

\paragraph{Percentile CI}
Let $\hat\theta_{n, \alpha}^{(b)}$ be the empirical $100\alpha$th percentile of
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$. The $(1 - \alpha)$
empirical percentile CI is
\[
(\hat\theta_{n, \alpha/2}^{(b)}, \hat\theta_{n, 1 - \alpha/2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$.
As will be shown in our simulation study, this approach works well for the
marginal mean and standard deviation of a serially dependent process, but its
coverage of the temporal dependence deteriorates as $n$ increases, which is in
contrary to what one would expect.

\paragraph{Centered Percentile CI}
Centered percentile CI is centered at the point estimator and uses the
variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\delta$, the difference between a point estimate and
$\bar\theta_n^{(b)}$. For $b \in \{1, \ldots, B\}$, we can obtain a
$\delta^{(b)}$ based on the $b$th bootstrap point estimate. Let
$\delta_{\alpha}^{(b)}$ be the empirical $100\alpha$th percentile of
$\{\delta^{(1)}, \ldots,  \delta^{(B)}\}$.  
A $(1 - \alpha)$ centered percentile CI is
\[
(\hat{\theta}_{n} + \delta^{(b)}_{\alpha/2}, \quad
\hat{\theta}_{n} + \delta^{(b)}_{1 - \alpha/2}).
\]
This CI is necessarily centered by the point estimator $\hat\theta_n$ and is
asymmetric, as different critical values are used to compute
the lower and upper bounds.

\paragraph{Bias Corrected and Accelerated (BCA) Bootstrap CI}
In this study, the BCA CI is an interval centered at the point estimator.
\jy{Isn't this just the empirical CDF? We just need the sample quantile functon.
  What is your reference for this description?}
Let $\hat{G}(c) = \#\{\hat\theta_n^{(b)} < c\}/B$ for $b \in \{1, \ldots, B\}$, let
$z_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$ for $b \in \{1, \ldots, B\}$, let
$\hat{\theta}_{n, -i}$ be the statistic of the sample without the $i^{th}$
block, and let $a$ be $SKEW\{\hat{\theta}_{n, -i}\} / 6$. Then
 $\hat\theta_{BCA}[\alpha] = \hat{G}^{-1}\Phi(z_0 + \frac{z_{0} + z_{\alpha}}{1 - a(z_{0} + z_{\alpha})}).$
A $(1 - \alpha)$ centered percentile CI is
\[
(\hat\theta_{BCA}[\alpha/2], \quad
\hat\theta_{BCA}[1 - \alpha/2]).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$.
\jy{why is it called BCA?}

@@

We designed a simulation study to compare the performance of the different
block
bootstrap CIs. In particular, we generated time series $X_t$
from a 1st order auto-regressive (AR1) process:
\[
X_t = \phi X_{t-1} + \epsilon_t,
\]
where $\phi$ is an auto-regressive coefficient, and $\epsilon_t$ is a series of
independent errors from a normal distribution with mean zero and variance
$\sigma_{\epsilon}^2$. The strength of the serial dependence is controlled by
$\phi$, which was set to five levels: $\{-0.4, -0.2, 0.0, 0.2, 0.4\}$.
The series $X_t$ has mean zero and variance
$\sigma_x^2 = (1 - \phi^2) \sigma_{\epsilon}^2$, so for each value of $\phi$,
we
set $\sigma_{\epsilon}^2 = 1 / (1 - \phi^2)$ such that $\sigma_x^2 = 1$.

@@

\jy{Make it flow better and conciser}
The coverage rates of the CIs were used to compare the performance of CIs. Let
$\hat\theta_{L, CI}$ and $\hat\theta_{U, CI}$ be the lower bound and upper bound,
respectively,  for $CI \in \{1, \ldots, n_{CI}\}$, where $n_{CI}$ is the number of 
CIs.
Then $cov = \#\{\hat\theta_{L, CI} < \theta < \hat\theta_{U, CI} \}/n_{CI}$,  
for $CI \in \{1, \ldots, n_{CI}\}$.
The coverage rate is, of course, a
point estimate of a proportion, so a 95\% CI of this coverage
proportion (with $n_{CI} = 10000$) can be constructed. If the proportion
.95 is included in the interval, the block bootstrap method is likely
performing
well. If all values in the interval are below .95, the results suggest that the
method either is providing an inaccurate estimation, is underestimating the
process' variability, or a combination of both. If all values in the interval
are above .95, the results suggest that the method is overestimating the
process' variability.

@@

The results are graphically displayed in Figures~\ref{fig:mu}--\ref{fig:phi};
these figures were generated using the using
the R ggplot2 package \citep{ggplot2}.

@@

Based on \ref{fig:mu}, performance is
about the same for the three interval types studied for $\mu$. As dependence gets stronger,
a larger sample size is necessary. For $\mu$, all three types are working asymptotically 
but for finite sizes, they more or less need about the same sample size to have the 
correct coverage. When $\phi < 0$,
$n = 100$ seems to  be large enough to cover $\mu$ at the nominal rate. When
$\phi = 0$, or in other other words, the sample is from a normal distribution,
$n = 500$ seems to be large enough to cover $\mu$ at the nominal rate. Samples
from processes with $\phi > 0$ appear to require a larger $n$ than samples from
processes with $\phi < 0$. When $\phi = 0.2$, $n = 300$ seemed to be necessary
for adequate performance, and when $\phi = 0.4$, an even larger $n$ appeared
to be necessary to properly cover $\mu$. It is rare for $\phi$ to be so high,
so in most cases, if $\phi$ is unknown, a sample of $n \geq 500$ should be
used when the target parameter is $\mu$.

>>>>

\bigskip
\noindent{\sc Keywords}:
dependent data; resampling; simulation; time series
\end{abstract}

@@

The necessary sample size for bootstrap standard errors to provide valid
uncertainty measures in practice has not been extensively studied. For
independent data and non-block bootstrap, \citet{hesterberg2015teachers} notes
that while percentile-based CIs from nonparametric bootstrap
are more accurate than $t$-intervals for larger sample sizes, they are
less accurate for smaller sample sizes. In the context of structural equation
modeling, \citet{nevitt2001performance} find
that a sample size of 200--1000 is usually sufficient for interval estimation
using standard nonparametric bootstrap. For dependent data, the necessary sample
size for block bootstrap is even less clear. In a linear regression setting with
dependent data, \citet{goncalves2005bootstrap} find that standard error
estimates from block bootstrap in small samples may be more accurate than
inference from closed-form asymptotic estimates, but block bootstrap percentile
CIs still do not have sufficient coverage even for sample size
$n = 1024$.

@@

The remainder of the paper is organized as follows.
Section~\ref{sec:bbci} reviews block bootstrap procedures and how to use block
bootstrap estimates to construct CIs. Section~\ref{sec:simu} reports a simulation
study comparing the coverage rates of the block bootstrap CIs. A discussion
concludes in Section~\ref{sec:disc}.

@@

Consider a stationary time series $\{X_t: t = 1, \ldots, n\}$ with length~$n$.
Our goal is to construct a CI for a parameter $\theta$ in the
data generating model of the series. Suppose that $\hat\theta_n$ is a point
estimator of~$\theta$ based on the observed series. Bootstrap is a powerful
approach to construct CIs. If the observations in the series
were independent, a standard nonparametric bootstrap procedure would draw a
large number~$B$ bootstrap copies of the observed data, and calculate an
estimate $\hat\theta_n^{(b)}$ for each copy $b = 1, \ldots, B$. The uncertainty
of $\hat\theta_n$ is then estimated by the empirical uncertainty of the
$\hat\theta_n^{(b)}$'s. When serial dependence is present, the bootstrap
procedure needs to preserve the serial dependence. Block bootstrap was
motivated for this situation.

@@

Block bootstrap preserves the serial dependence in the observed data by
partitioning the data into blocks and performing bootstrap on the blocks.
In particular, consider block size~$l$ and, for convenience, suppose that
$n$ is a multiple of $l$ such that there are $k = n / l$ blocks. Each block~$j$
is $Y_j = \{X_{(j - 1) * l + 1}, \ldots, X_{(j - 1) * l + l}\}$,
$j = 1, \ldots,   k$.  Then, we sample $k$ blocks of $Y_j$'s from the set 
$\{Y_1, \ldots, Y_k\}$ with replacement and concatenate the $k$ sampled blocks
in the order they are picked to form a bootstrap sample of the data. The
formation of the bootstrap sample ensures that the between-block dependence is
weak and that the within-block serial dependence is preserved. Because the
blocks here are non-overlapping, this bootstrap approach is known as
non-overlapping block bootstrap, or simple block bootstrap.

Alternatively, block-bootstrap can be done with overlapping or moving blocks.
Define moving blocks $Z_j = \{X_j, \ldots, X_{j + l - 1}\}$,
$j = 1, \ldots, n - l + 1$. Now we draw $k$ blocks from the $(n - l + 1)$
blocks
of $Z_j$'s with replacement and then align them in the order they were
picked
to form a block bootstrap sample. If $n$ is not a multiple of~$l$, the last
block selected will be reduced in size so that the final size of the
block bootstrap sample is $n$. It is also possible to implement moving block
bootstrap while allowing blocks to wrap around the end of the series. In other
words, define moving blocks (assuming $l > 1$):
\jy{Any equation must be a part of a sentence and should be punctuated.}
\jy{Recall a blank line means a new paragraph.}
\jy{Define your response command with color and respond}

@@

Suppose that we have repeated the steps in the last subsection $B$ times, and
that for $b \in \{1, \ldots, B\}$, we have obtained an estimator
$\hat\theta_n^{(b)}$ based on the $b$th bootstrap sample using the same method
that was applied to $\{X_t: t = 1, \ldots, n\}$ to obtain $\hat\theta_n$.
Now the question is, how to construct a CI for $\theta$
using the $B$ bootstrap point estimates
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$.
We consider three kinds of CIs.

@@

\paragraph{Percentile CI}
Let $\hat\theta_{n, \alpha}^{(b)}$ be the empirical $100\alpha$th percentile of
$\{\hat\theta_n^{(1)}, \ldots, \hat\theta_n^{(B)}\}$. The $(1 - \alpha)$
empirical percentile CI is
\[
(\hat\theta_{n, \alpha/2}^{(b)}, \hat\theta_{n, 1 - \alpha/2}^{(b)}).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$.
As will be shown in our simulation study, this approach works well for the
marginal mean and standard deviation of a serially dependent process, but its
coverage of the temporal dependence deteriorates as $n$ increases, which is
contrary to what one would expect.

\paragraph{Centered Percentile CI}
Centered percentile CI is centered at the point estimator and uses the
variation
in the bootstrap estimates to construct the error bound. This interval requires
the computation of $\bar\theta_n^{(b)}$, the mean of all bootstrap point
estimates, and $\delta$, the difference between a point estimate and
$\bar\theta_n^{(b)}$. For $b \in \{1, \ldots, B\}$, we can obtain a
$\delta^{(b)}$ based on the $b$th bootstrap point estimate. Let
$\delta_{\alpha}^{(b)}$ be the empirical $100\alpha$th percentile of
$\{\delta^{(1)}, \ldots,  \delta^{(B)}\}$.  
A $(1 - \alpha)$ centered percentile CI is
\[
(\hat{\theta}_{n} + \delta^{(b)}_{\alpha/2}, \quad
\hat{\theta}_{n} + \delta^{(b)}_{1 - \alpha/2}).
\]
This CI is necessarily centered around the point estimator $\hat\theta_n$ and is
asymmetric, as different critical values are used to compute
the lower and upper bounds.

\paragraph{Bias Corrected and Accelerated (BCA) Bootstrap CI}
In this study, the BCA CI is an interval centered at the point estimator.
\eds{a different abbrevation/notation was used for this in the introduction}
\jy{Isn't this just the empirical CDF? We just need the sample quantile functon.
  What is your reference for this description?}
Let $\hat{G}(c) = \#\{\hat\theta_n^{(b)} < c\}/B$ for $b \in \{1, \ldots, B\}$, let
$z_0 = \Phi^{-1}\{\#\{\hat\theta_n^{(b)} < \hat{\theta}_n\} / B\}$ for $b \in \{1, \ldots, B\}$, let
$\hat{\theta}_{n, -i}$ be the statistic of the sample without the $i^{th}$
block, and let $a$ be $SKEW\{\hat{\theta}_{n, -i}\} / 6$. Then
 $\hat\theta_{BCA}[\alpha] = \hat{G}^{-1}\Phi(z_0 + \frac{z_{0} + z_{\alpha}}{1 - a(z_{0} + z_{\alpha})}).$ \eds{notation will need to be cleaned and clarified}
A $(1 - \alpha)$ centered percentile \eds{BCA?} CI is
\[
(\hat\theta_{BCA}[\alpha/2], \quad
\hat\theta_{BCA}[1 - \alpha/2]).
\]
This CI is not necessarily centered by the point estimator $\hat\theta_n$.
\jy{why is it called BCA?}

@@

We designed a simulation study to compare the performance of the different
block
bootstrap CI methods. In particular, we generated time series $X_t$
from a 1st order auto-regressive (AR1) process:
\[
X_t = \phi X_{t-1} + \epsilon_t,
\]
where $\phi$ is an auto-regressive coefficient, and $\epsilon_t$ is a series of
independent errors from a normal distribution with mean zero and variance
$\sigma_{\epsilon}^2$. The strength of the serial dependence is controlled by
$\phi$, which was set to five levels: $\{-0.4, -0.2, 0.0, 0.2, 0.4\}$.
The series $X_t$ has mean zero and variance
$\sigma_x^2 = (1 - \phi^2) \sigma_{\epsilon}^2$, so for each value of $\phi$,
we
set $\sigma_{\epsilon}^2 = 1 / (1 - \phi^2)$ such that $\sigma_x^2 = 1$.

@@

\jy{Make it flow better and conciser}
The coverage rates of the CIs were used to compare the performance of CIs. Let
$\hat\theta_{L, CI}$ and $\hat\theta_{U, CI}$ be the lower bound and upper bound,
respectively,  for $CI \in \{1, \ldots, n_{CI}\}$, where $n_{CI}$ is the number of 
CIs.
Then $cov = \#\{\hat\theta_{L, CI} < \theta < \hat\theta_{U, CI} \}/n_{CI}$,  
for $CI \in \{1, \ldots, n_{CI}\}$.
The coverage rate is, of course, a
point estimate of a proportion, so a 95\% CI of this coverage
proportion (with $n_{CI} = 10000$) can be constructed. If the proportion
.95 is included in the interval, the block bootstrap method is likely
performing
well. If all values in the interval are below .95, the results would suggest 
that the
method either is providing inaccurate estimation, is underestimating the
process' variability, or a combination of both. If all values in the interval
are above .95, the results suggest that the method is overestimating the
process' variability.

@@

The results are graphically displayed in Figures~\ref{fig:mu}--\ref{fig:phi};
these figures were generated using the R \textsl{ggplot2} package \citep{ggplot2}.

@@

Based on Figure~\ref{fig:mu}, the performance is
about the same for the three interval types studied for $\mu$. As dependence 
gets stronger,
a larger sample size is necessary. For $\mu$, all three types are working asymptotically 
but for finite sizes, they more or less need about the same sample size to have 
the correct coverage. When $\phi < 0$,
$n = 100$ seems to  be large enough to cover $\mu$ at the nominal rate. When
$\phi = 0$, or in other other words, the sample is independently normally
distributed, %from a normal distribution,
$n = 500$ \eds{$n=400$?} seems to be large enough to cover $\mu$ at the nominal 
rate. Samples
from processes with $\phi > 0$ appear to require a larger $n$ than samples from
processes with $\phi < 0$. When $\phi = 0.2$, $n = 300$ seemed to be necessary
for adequate performance, and when $\phi = 0.4$, an even larger $n$ appeared
to be necessary to properly cover $\mu$. It is rare for $\phi$ to be so high,
so in most cases, if $\phi$ is unknown, a sample of $n \geq 500$ should be
used when the target parameter is $\mu$. \eds{I would argue that none of the 
sample sizes considered work well for $\phi=0.2$ or $\phi=0.2$ based on 
Figure~\ref{fig:mu}}
